\subsection{PVS}

When a PVS file is regenerated, the old version is split into chunks,
according to blank lines. Chunks corresponding to \why declarations
are identified with a comment starting with \verb+% Why3+, \eg
\begin{verbatim}
  % Why3 f
  f(x: int) : int
\end{verbatim}
Other chunks are considered to be user PVS declarations.
Thus a comment such as \verb+% Why3 f+ must not be removed;
otherwise, there will be two
declarations for \texttt{f} in the next version of the file (one being
regenerated and another one considered to be a user-edited chunk).

The user is allowed to perform the following actions on a PVS
realization:
\begin{itemize}
\item give a definition to an uninterpreted symbol (type, function, or
  predicate symbol), by adding an equal sign (\texttt{=}) and a
  right-hand side to the definition. When the declaration is
  regenerated, the left-hand side is updated and the right-hand side
  is reprinted as is. In particular, the names of a function or
  predicate arguments should not be modified. In addition, the
  \texttt{MACRO} keyword may be inserted and it will be kept in
  further generations.

\item turn an axiom into a lemma, that is to replace the PVS keyword
  \texttt{AXIOM} with either \texttt{LEMMA} or \texttt{THEOREM}.

\item insert anything between generated declarations, such as a lemma,
  an extra definition for the purpose of a proof, etc.
\end{itemize}
\why makes some effort to merge the new declarations with the old ones
and with the user chunks. If it happens that some chunks could not be
merged, they are appended at the end of the file, in comments.
